Finiteness in Cubical Type Theory

### 2020-05-27 paper.tex
Update paper.tex

\section{Introduction}
\subsection{Foreword}
In constructive mathematics we are often preoccupied with \emph{why} something
is true.
Take finiteness, for example.
There are a handful of ways to demonstrate some type is finite: we could provide
a surjection to it from another finite type; we could show that any collection
of its elements larger than some bound contains duplicates; or we could show
that any stream of its elements contain duplicates.

Classically, all of these proofs end up proving the same thing: that our type is
finite.
Constructively (in Martin-Löf Type Theory (MLTT)
\cite{martin-lofIntuitionisticTypeTheory1980} at least), however, all three
of the statements above construct a different version of finiteness.
\emph{How} we show that some type is finite has a significant impact on the
type of finiteness we end up dealing with.

Homotopy Type Theory (HoTT) \cite{hottbook} adds another wrinkle to the story.
Firstly, in HoTT we cannot assume that every type is a (homotopy) set: this
means that the finiteness predicates above can be further split into versions
which apply to sets only, and those that apply to all types.
Secondly, HoTT gives us a principled and powerful way to construct quotients,
allowing us to regain some of the flexibility of classical definitions by
``forgetting'' the parts of a proof we would be forced to remember in MLTT.

Finally, for a computer scientist constructive mathematics has one invaluable
feature missing from classical mathematics: computation.
Cubical Type Theory (CuTT) \cite{cohenCubicalTypeTheory2016}, and its
implementation in Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, realise
this property even in the presence of univalence, giving computational content
to programs written in HoTT.
\subsection{Contributions}
The finiteness predicates we are interested in are organised in
figure~\ref{finite-classification}, and defined in
section~\ref{finiteness-predicates}.
We will explore two aspects of each predicate: its relation to the other
predicates, and its topos-theoretic classification.

>>>>

\section{Introduction}
\subsection{Foreword}
Proofs in constructive mathematics are often more substantial than their
classical counterparts.
``Substantial'' here doesn't mean the literal length of the proofs (although the
double meaning is not lost on anyone who has had to struggle with a modern
constructive proof assistant), but rather refers to their \emph{content}: a
constructive proof contains more \emph{stuff} than a classical one.

If we were to prove that a set was finite in classical mathematics, we might
show that it is in bijection with another finite set, say a finite prefix of the
natural numbers.
In semi-formal notation, to say that a set \(A\) is finite we would show the
following:
\begin{equation} \label{example-finite}
  \mathbf{Finite}(A) \coloneqq \exists (n : \mathbb{N}) . \; A \Leftrightarrow \left\{ m \mid m < n\right\}
\end{equation}
Where \(A~\Leftrightarrow~B\) means ``a bijection between \(A\) and \(B\)''.

Classically, a proof of equation~\ref{example-finite} is nothing more than
\emph{evidence} for the proposition of equation~\ref{example-finite}.
A constructive proof of \ref{example-finite}, however, could include much more:
\begin{enumerate}
  \item A number \(n\) representing the cardinality of \(A\).
  \item A function from \(A\) to \(\left\{ m \mid m < n \right\}\).
  \item A function from \(\left\{ m \mid m < n \right\}\) to \(A\).
\end{enumerate}
In other words, it makes sense (in constructive mathematics) to talk about a
function of type \(\mathbf{Finite}(A)~\rightarrow~\mathbb{N}\), which takes a
\emph{proof} of the finiteness of \(A\) and returns \(A\)'s cardinality.

Such functions are perhaps even more concrete than this description suggests:
many constructive proofs are in fact computer programs in disguise (and
vice-versa \cite{wadlerPropositionsTypes2015}).
All ``proofs'' in this paper correspond to honest-to-goodness proofs in a
programming language (Cubical Agda \cite{vezzosiCubicalAgdaDependently2019} in
our case) which can be run on a computer to produce real output.

We have now hopefully established, albeit informally, what it means for
constructive proofs to be ``substantial''.
Clearly this extra information carried by proofs can add richness to their
study: we will see, for instance, no fewer than five finiteness predicates in
this paper which would collapse to one in a classical setting.
The extra information also presents a challenge, however: occasionally we will
see that we have revealed \emph{too much} information while proving some
property.
For instance, in order to construct a category of finite sets we will need to
describe the objects of that category: constructively this would be ``pairs of
sets along with proofs that they are finite''.
The category we're trying to construct has one object for every finite set (up
to unique isomorphism, etc).
Here's the problem: if we take our finiteness proof to be a naive translation of
the predicate in equation~\ref{example-finite}, then the proof includes a
permutation on the set it refers to.
This means that two objects \((A, \mathbf{Finite}(A))\) and \((A,
\mathbf{Finite}(A))\) need not be the same, as they could have \emph{different
  permutations} embedded in the \(\mathbf{Finite}(A)\) term.
Where we wanted one object per set, we now have \(n!\) objects for every set of
cardinality \(n\).

Being able to hide the extra information in a constructive proof when we need to
is clearly important, then.
HoTT (and CuTT by extension) facilitate this hiding: how exactly it is
accomplished will be explored in the subsequent sections.

Finally, for all of its benefits, constructive mathematics (especially
dependently-typed constructive mathematics on a computer) still suffers from
many painful drawbacks.
As already hinted at, proofs in a constructive setting can tend to be verbose,
especially when the arbiter of correctness is as strict as a computer, which
neither understands nor trusts the time-honoured technique of ``the other cases
are obviously true''.
While the goal of using a computer to automate the boring parts of a proof still
seems attainable, too often getting a computer involved in a proof means adding
thousands of lines of tedium to satisfy its lack of intuition.

Simple proof search over a finite domain is a simple, if effective, way to
relieve some of the tedium.
Many libraries in Agda, Coq, and other languages make use of this fact.
While the idea for finite proof search is simple, we found that implementing one
revealed a number of interesting theoretical considerations:
\begin{description}
  \item[How should we construct search spaces?]
    We found that a \(Pi\)-pretopos provided the appropriate ``generator
    language'' for constructing search spaces for finite types.
  \item[How do we deal with functions?]
    We of course would like to be able to search over functions: \todo{?}
  \item[How do we talk about ``searchability''?]
    \todo{Omniscience}
\end{description}
\subsection{Contributions}
The finiteness predicates we are interested in are organised in
figure~\ref{finite-classification}, and defined in
section~\ref{finiteness-predicates}.
We will explore two aspects of each predicate: its relation to the other
predicates, and its topos-theoretic classification.
